Efficient Constructions of Prefer-Max Sequences and of Infinite de Bruijn Sequences

### 2017-01-05 paper.tex
title change
Yotam

\title{An Efficient Construction of the Prefer-Max Sequence and of an Infinite de Bruijn Sequence}

@@

A $k$-ary de Bruijn sequence of span $n$ is a cyclic sequence of length $k^n$ in which each $k$-ary string of length $n$ appears exactly once as
a substring. A well known construction of such a sequence was first proposed by Ford~\cite{Ford1957} as follows:

\begin{algorithm}
Repeatedly append the maximal symbol such that the word formed by the $n$ most recent symbols is new (in the first $n$ steps, act as if thee are $n$ invisible zeros before the sequence). Stop after appending $n$ consecutive zeros. 
\caption{The prefer-max sequence.}
\end{algorithm}

For example, if we choose $k=3$ and $n=3$ we get the sequence:
$$\langle2,2,2,1,2,2,0,2,1,1,2,1,0,2,0,1,2,0,0,1,1,1,0,1,0,0,0\rangle.$$
Note that this sequence begins with a prefix (the first $19$ symbols) that contains all the words in $\{0,1,2\}^n \setminus \{0,1\}^n$  (the words that contain the symbol $2$) as substrings. Note also that the rest of the sequence, after this prefix, is exactly the sequence that we would have obtained if we ran the construction with the alphabet $k=2$ (and the same $n$). The first result of this paper, that we call the Onion lemma, is that this is not a coincidence:

\begin{theorem}[Onion]
	For any $k,n \in \N$, the prefix of length $k^n-(k-1)^n$ of the sequence generated by the prefer-max construction is such that its $n$-length substrings are exactly $\{0,\dots,k\}^n \setminus \{0,\dots,k-1\}^n$, i.e., all the words of length $n$ that contain the symbol $k$. Furthermore, the suffix that begins with the 
	$(k^n-(k-1)^n+1)$th symbols is exactly the sequence generated by the prefer-max construction with the parameters $k-1$ and $n$.
	\label{thm:onion}
\end{theorem}

This observation lead us to look at the reverse of the prefer-max sequence. Specifically, the structure identified in Theorem~\ref{thm:onion} means that if we somehow manage to construct the sequence backwards, we do not have to specify $k$ ahead of time because we can go on indefenitiley while $k$ grows in layers: we first write $0^n$, then a word whose substrings of length $n$ are $\{0,1\}^n$, then extend with other words in $\{0,1,2\}^n$, and so on.

>>>>

\title{Efficient Constructions of Prefer-Max Sequences and of Infinite de Bruijn Sequences}

@@

A $k$-ary de Bruijn sequence of span $n$ is a cyclic sequence of length $k^n$ in which each $k$-ary string of length $n$ appears exactly once as
a substring. A well known construction of such a sequence, first proposed by Ford~\cite{Ford1957}, can be described as follows:

\begin{algorithm}
Repeatedly append the maximal symbol over the alphabet $0,\dots,k-1$ such that the word formed by the $n$ most recent symbols is new (in the first $n$ steps, act as if thee are $n$ invisible zeros before the sequence). Stop after appending $n$ consecutive zeros. 
\caption{The $(k,n)$-prefer-max sequence.}
\end{algorithm}

For example, if we choose $k=3$ and $n=3$ we get the sequence:
$$\langle2,2,2,1,2,2,0,2,1,1,2,1,0,2,0,1,2,0,0,1,1,1,0,1,0,0,0\rangle.$$

An obvious limitation of the above construction is, of course, that, when $n$ and $k$ grow, the amount of time and memory needed to produce the next symbol grows exponentially. Therefore, this construction is not practical for generating de Bruijn sequences of large alphabets and spans. It is used here as a mathematical definition of the sequence. An efficient algorithm for constructing this sequence is one of the contributions of this paper.

Note that the $(n,k)$-prefer-max sequence begins with a prefix (the first $19$ symbols) that contains all the words in $\{0,1,2\}^n \setminus \{0,1\}^n$  (the words that contain the symbol $2$) as substrings. Note also that the rest of the sequence, after this prefix, is exactly the sequence that we would have obtained if we ran the construction with the alphabet $k=2$ (and the same $n$). The first result of this paper, that we call the Onion lemma, is that this is not a coincidence:

\begin{theorem}[Onion]
	The last $(k-1)^n+1$ symbols of the $(k,n)$-prefer-max sequence form the $(k-1,n)$-prefer-max sequence.
\end{theorem}
\begin{proof}
	Denote the symbols of the $(n-k$)-prefer-max sequence by $s_1,s_2,\dots,s_{k^n+1}$ and let, for notational convenience, $s_i=0$ for all $i<1$. Let ${i_0} = \min\bigl\{i\colon (k-1) \notin \{s_{i-n+1},\dots,s_{i}\}\bigr\}$. 
	By minimality we have that $s_{{i_0}-n}=k-1$. Since $s_{i_0} \neq k-1$ and because $s_{{i_0}-n+1}\cdots s_{{i_0}-1} (k-2)$ did not appear as a substring before ${i_0}$ we get, by construction, that $s_{{i_0}}=k-2$. This means that the exists some ${i_1}<{i_0}$ such that $s_{{i_1}-n+1}\cdots s_{{i_1}} = s_{{i_0}-n+1}\cdots s_{{i_0}-1}(k-1)$.

	Assume, towards contradiction, that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} \neq 0^{n-1}$. In particular $s_{{i_1}-n+1}\cdots s_{{i_1}} \neq 0^{n-1}(k-1)$ which means that ${i_1} > 1$. Therefore, $k-1  \in \{s_{{i_1}-n},\dots,s_{{i_1}-1}\}=\{s_{{i_1}-n}\} \cup \{s_{{i_0}-n+1},\dots,s_{{i_0}-1}\}$. Since the second set in this union does not contain $k-1$, we get that $s_{{i_1}-n}=k-1$. This leads to a contradiction because it means that $s_{{i_1}-n}\cdots s_{{i_1}-1} = s_{{i_0}-n}\cdots s_{{i_0}-1}$ which cannot happen in a de Bruijn sequence. This contradiction gives us that $s_{{i_0}-n+1}\cdots s_{{i_0}-1} = 0^{n-1}$.

	The sequence is therefore as follows:
	$$\underbrace{s_1,\dots,s_{i_0-n-1},k-1,{0,\dots,0}}_{\parbox{10 em}{\scriptsize \centering{all substrings of length $n$ contain $k-1$}}},\underbrace{k-2}_{s_{i_0}},s_{i_0+1},\dots,s_{k^n+1}$$
	To complete the proof, we need to show that all substrings of length $n$ that contain $k-1$ appear before $s_{i_0}$.

	For $w \in \{0,\dots,k-1\}^n$ and $1 \leq m \leq n$, let $j_m$ be such that $s_{j_m-n+1}\cdots s_{j_m} = w[m..n]0^{m-1}$, where $w[m..n]$ is the $n-m$ letters suffix of $w$. 
	Because $0$ is the smallest symbol in our alphabet, we get that it is added only if all the other symbols cannot be added, i.e., when adding other symbols would generate a substring that has already being seen before. 
	%In particular, we get that $w[2..m]0$ appears as a substring in the $(n,k)$-prefer-max sequence before $w$, 
	Since $w$ comes right before some $w[2..m]\tau$ as a window, since $\tau \geq 0$,  $w[2..m]0$ comes after (or is equal to) $w[2..m]\tau$ which is after $w$. Similarly we can see that $w[2..m]0$ appears before $w[3..m]0^2$ and so on.
	This gives us that $j_1 < j_2 < \cdots < j_n$. Note that $s_{j_1-n+1}\cdots s_{j_1}=w$, i.e, that $w$ appears as a substring at the $j_1$'s symbol of the sequence.

	We will now show that if $w$ contains the symbol $k-1$ it must appear as a substring before the $i_0$'s symbol in the sequence. First consider the case where the last (right-most) symbol of $w$ is $k-1$. In this case, $j_n=i_{0}-1$ so we have, because $j_1 < j_n$, that $w$ appears as a substring before the $i_0$'th symbol in the sequence. The second case is when the last (right-most) symbol in $w$ is not $k-1$ but there is another letter in $w$ which is $k-1$.

*************

%	$i^*$ %=\idx{\{0,\dots,k-2\}^n, 1}$ 
%	be the first index such that $$of the first tuple in the $(k,n)$-prefer-max sequence that does not contains $k-1$.

	Clearly, $s_{i^*}=\T{\xi_1,\dots,\xi_{n-1},k-2}$ and $s_{i^*-1}=\T{k-1,\xi_1,\dots,\xi_{n-1}}$ for some $\T{\xi_1,\dots,\xi_{n-1}}\in\{0,\dots,k-2\}^{n-1}$. By Observation~\ref{claim:x0 is last}, there is some $i<i^*$ such that $s_i=\T{\xi_1,\dots,\xi_{n-1},k-1}$. If $\T{\xi_1,\dots,\xi_{n-1}} \neq 0$ then $i \neq 1$ so we can examine the tuple $s_{i-1}$. By the minimality of $i^*$, the tuple $s_{i-1}$ must contain $k-1$ therefor, because $\T{\xi_1,\dots,\xi_{n-1}}\in\{0,\dots,k-2\}^{n-1}$, we must have $s_{i-1}=\T{k-1,\xi_1,\dots,\xi_{n-1}}$ which is a contradiction because it means that $s_{i-1} = s_{i^*-1}$ for some $i<i^*$. To avoid this contradiction we must have that  $s_i*=\T{0,\dots,0,k-2}$ and that $s_{i*-1}=\T{k-1,0,\dots,0}$

	We are left to show that all the tuples that contain $k-1$ appear before $s_{i*-1}=\T{k-1,0,\dots,0}$.
	By Observation~\ref{claim:0 follower comes after}, for all $\T{\xi_1,\dots,\xi_{n-1}}$ we have that
	\begin{eqnarray*}
		\idx(\T{k-1,\overbrace{0,\dots ,0}^{n-1 \text{ times}}}) 
		& > & \idx(\T{\xi_{n-1},k-1,\overbrace{0,\dots ,0}^{n-2 \text{ times}}}) \\
		& > & \idx(\T{\xi_{n-2},x_{n-1},\xi,\overbrace{0,\dots ,0}^{n-3 \text{ times}}}) \\    
		& \vdots \\
		& > &  \idx(\T{\xi_1,\dots,\xi_{n-1},k-1})
	\end{eqnarray*}
	This establishes that every tuple that ends with a $k-1$ at the right comes before $\T{k-1,0,\dots ,0}$. To see that this is also true for any tuple that contain $k-1$ at any of its coordinates, note that for such a tuple there is a tuple that comes at most $n$ indexes before it that has $k-1$ on the right. To complete the proof note also that all the tuples between these two tuples have $k-1$ at some coordinate that is not first from the left so we know that $\T{k-1,0,\dots ,0}$ does not come between them. 
\end{proof}

%\begin{theorem}[Onion]
%	For any $k,n \in \N$, the prefix of length $k^n-(k-1)^n$ of the sequence generated by the prefer-max construction is such that its $n$-length substrings are exactly $\{0,\dots,k\}^n \setminus \{0,\dots,k-1\}^n$, i.e., all the words of length $n$ that contain the symbol $k$. Furthermore, the suffix that begins with the 
%	$(k^n-(k-1)^n+1)$th symbols is exactly the sequence generated by the prefer-max construction with the parameters $k-1$ and $n$.
%	\label{thm:onion}
%\end{theorem}
%\begin{proof}
%	
%\end{proof}

This observation lead us to examine the reverse of the prefer-max sequence. Specifically, the structure identified in Theorem~\ref{thm:onion} means that if we somehow manage to construct the sequence backwards, we do not have to specify $k$ ahead of time. We can add symbols forever while $k$ grows in layers: we first write $0^n$, then a word whose substrings of length $n$ are $\{0,1\}^n$, then extend with other words in $\{0,1,2\}^n$, and so on.
